Escape Analysis and Specialization in a Tracing JIT

### 2010-09-26 talk/pepm2011/paper.tex
reorder some stuff, write two paragraphs about pypy

The tracing JIT approach that the PyPy project is taking removes the overhead
of bytecode dispatch. In this paper we want to explain how the traces that are
produced by our meta-tracing JIT can be optimized to also remove some of the
overhead more closely associated to dynamic languages, such as boxing overhead
and type dispatching. To understand the problem more closely, we analyze the
occurring object lifetimes in Section~\ref{sec:lifetimes}. The most important
technique to achieve this is a form of escape analysis \cite{XXX} that we call
\emph{virtual objects}, which is described in Section~\ref{sec:virtuals}. The
basic approach of virtual objects can then be extended to also be used for
type-specializing the traces that are produced by the tracing JIT
(Section~\ref{sec:crossloop}). In Section~\ref{sec:XXX} we describe some
supporting techniques that are not central to the approach, but are needed to
improve the results. The introduced techniques are evaluated in
Section~\ref{sec:Evaluation}.

@@

\begin{enumerate}
    \item An efficient and effective escape analysis for a tracing JIT
    \item XXX
\end{enumerate}

@@

>>>>

The tracing JIT approach that the PyPy project is taking removes the overhead
of bytecode dispatch. In this paper we want to explain how the traces that are
produced by PyPy's tracing JIT can be optimized to also remove some of the
overhead more closely associated to dynamic languages, such as boxing overhead
and type dispatching. To understand the problem more closely, we analyze the
occurring object lifetimes in Section~\ref{sec:lifetimes}. The most important
technique to achieve this is a form of escape analysis \cite{XXX} that we call
\emph{virtual objects}, which is described in Section~\ref{sec:virtuals}. The
goal of virtual objects is to remove allocations of temporary objects that have
a predictable lifetime and to optimize type dispatching in the process.

The basic approach of virtual objects can then be extended to also be used for
type-specializing the traces that are produced by the tracing JIT
(Section~\ref{sec:crossloop}). In Section~\ref{sec:XXX} we describe some
supporting techniques that are not central to the approach, but are needed to
improve the results. The introduced techniques are evaluated in
Section~\ref{sec:Evaluation} using PyPy's Python interpreter as a case study.

@@

\begin{enumerate}
    \item An efficient and effective algorithm for removing objects allocations in a tracing JIT.
    \item XXX
\end{enumerate}

@@

The work described in this paper was done in the context of the PyPy project
\cite{armin_rigo_pypys_2006}. PyPy is an environment where dynamic languages can
be implemented in a simple yet efficient way. The approach taken when
implementing a language with PyPy is to write an interpreter for the language in
\emph{RPython} \cite{davide_ancona_rpython:_2007}. RPython ("restricted Python")
is a subset of Python chosen in such a way, that type inference becomes
possible. The language interpreter can thus be translated with the help of
PyPy's tools into a VM on the C level. Because the interpreter is written at a
relatively high level, the language implementation is kept free of low-level
details, such as object layout, garbage collection or memory model. Those
aspects of the final VM are woven into the generated code during the translation
to C.

The feature that makes PyPy more than a compiler with a runtime system is it's
support for automated JIT compiler generation \cite{bolz_tracing_2009}. During
the translation to C, PyPy's tools can generate a just-in-time compiler for the
language that the interpreter is implementing. This process is not fully
automatic, but needs to be guided by the language implementer by some
source-code hints.
